\textit{Supplemental materials for}: How is experience transformed into memory?

### 2018-09-04 paper/supplemental_materials.tex
backing up edits to supplemental materials [WIP]

\section*{Additional details about topic modeling methods and results}
\subsection*{Optimizing topic model parameters}
In order to create accurate video and recall models, we used an optimization method that was driven by our ability to explain hand-annotated memory performance metrics collected by \cite{ChenEtal17}.  Specifically, we used a grid search to compute the $\omega$ (movie sliding window duration, in scenes), $\rho$ (recall sliding window duration, in sentences), and $K$ (number of topics) that satisfied
\[
\argmax_{\omega, \rho, K} \mathrm{corr}\left(\mathrm{corr}\left(\mu\left(\omega, \rho, K\right), \nu\left(\omega, \rho, K\right)\right), \theta\right),
\]
where $\mathrm{corr}(\mu, \nu)$ is the per-participant correlation between the upper triangles of the temporal correlation matrices of the video ($\mu$) and recall ($\nu$) trajectory, and $\theta$ is the per-participant hand-annotated memory performance.  We searched over a grid of pre-specified values for each of these parameters; the resulting correlations are displayed in Figure~\ref{fig:paramsearch}.  The optimal parameters were $\omega = 250$, $\rho = 50$, and $K = 100$.

@@

\begin{figure}[tb]
\centering
\includegraphics[width=1\textwidth]{figs/feature_value}
\caption{\small \textbf{Feature importance analysis.} \textbf{A.} Contributions of each feature type to the structure of the video trajectory. The bar heights reflect the correlation between the video trajectory computed using all features with a video trajectory computed using all features except the indicated feature.  (Lower bars reflect features that contribute more substantially to the video trajectory's shape.) \textbf{B.} Which features are preserved during recall?  The bar heights reflect the (average) across-participant correlations between the video and recall trajectories.  Error bars denote 95\% confidence intervals.  \textbf{C.} Feature correlation matrix.  Each entry displays the correlation between video topic trajectories created using only the indicated (row/column) features.}
\label{fig:feature-importance}
\end{figure}

@@

\textbf{JRM STOPPED HERE.}

@@

\section*{Naturalistic extensions of classic list-learning analyses}
Just like in a traditional free recall list-learning experiment where participants view a list of words and then verbally recall them, our video-recall matching analysis approach affords us the ability to analyze memory in the same way. The recalled events can be treated as ``items'' analogous to words recalled in a list-learning study. Here, we sought to characterize memory performance/dynamics by extending classic analyses originally designed for list-learning experiments to more naturalistic settings.

First, we asked whether the estimated number of recall events ($k$) by participant was related to hand-annotated accuracy as published in Chen et al. (2017).  We found a strong positive correlation where participants with a greater number of recall events also had better overall memory performance (Pearson's $r(16) = 0.67, p = 0.003$). Then, we considered how participants initiated the recall sequence (known in the literature as the `probability of first recall' or `PFR'). We found that participants tended to initiate their recall sequences with the first few events (Supp. Fig.~\ref{fig:list-learning}A), which is qualitatively very similar to previously published list learning experiments ~\citep{HowaKaha99}. Next, we considered another well-studied memory measure in the list-learning literature, the lag conditional response probability curve (or lag-CRP) ~\citep{Kaha96}. The result suggests a strong bias to transition sequentially events in the forward direction (Supp. Fig.~\ref{fig:list-learning}B). Finally, we assessed memory performance for each event in the video as a function of its serial position during encoding (Supp. Fig.~\ref{fig:list-learning}C). We did not observe the classic ``primacy'' and ``recency'' pattern which is prevalent in the literature ~\citep{Murd62a}. We also considered two additional across-participant measures of recall that characterize memory organization: temporal clustering and semantic clustering. We found that participants who clustered in time also recalled a greater number of events (Pearson's $r(16) = 0.62, p = 0.007$). Next, we assessed semantic clustering. We found that the semantic clustering score was related to memory performance across participants (Pearson's $r(16) = 0.55, p = 0.02$).  Thus, participants who organized their recalls with respect to the semantic information contained in the scene had better memory performance.

\section*{Additional measures of naturalistic memory}
To quantify the similarity between the video model and individual recall models, we considered a number of novel metrics.  First, we tested whether each participant's recall model matched the video model in a general sense. To do this, for each participant we filtered the video model to only include the events that the participant remembered and computed the root mean squared difference (RMSD) between the video model and the recall model. As an example, if the participant remembered all the events in order (with perfect precision), the expected distance value would be 0. However, if they remembered a subset of events, events out of order, or with low precision, the expected distance would be greater than 0. To assess significance, we performed a permutation test where we circularly shifted the recall model (10000 times) and recomputed the RMSD. The recall model significantly matched the video model for nine of the participants ($p < 0.05$, participants: 3-4, 8-13, 17 and the p-value for the rest of the participants was less than .25). Furthermore, the RMSD values were significantly correlated to hand annotated memory performance across participants (Pearson's $r(16) =-.57, p = 0.016$). Thus, a closer match between the video and recall event models was related to better recall performance.

@@

\subsection*{List-learning analyses}
\textbf{Overall Accuracy.}  To get an overall measure of the quantity of information recalled, we computed the proportion of sucessfully recalled events by counting the number of unique recall events identified by the HMM model and dividing by the total number of video events.  We performed this analysis for each participant separately.

\textbf{Probability of first recall (PFR).}  The (PFR) analysis represents the probability that an item will be recalled first as a function of its serial position during encoding. We initialized a \# of participants (17) by \# of video events (34) matrix. Then for each participant, we found the index of the video event that was recalled first and filled in that index in the matrix with a 1.  Finally, we averaged over the rows of the matrix, resulting in a 1 by 34 array representing the proportion of participant that recalled an event as a function of serial position during encoding.

\textbf{Lag conditional probability curve (lag-CRP).} The lag-CRP represents the probability that the next item recalled will be of lag $i$ from the just recalled item. For each recall transition, we computed the lag between the current recall event and the next recall event, normalizing by the total number of possible transitions.  This resulted in a \# of participants (17) by lags (-33:+33) matrix. We averaged over the rows of this matrix to get a group-averaged lag-CRP.

\textbf{Serial position curve (SPC).} The SPC represents the proportion of participants that remember an item as a function of its serial position during encoding. We initialized a \# of participants (17) by \# of video events (34) matrix. Then, for each recall event (and each participant), we found the index of the video event that was recalled and filled it in with a 1. This resulted in a matrix where 1s indicate the successful recall of an event in serial position $n$ and zeros indicate the lack of recall for that event.  Lastly, we averaged over the rows of the matrix to get a 1 by 34 array representing the proportion of participants that recalled an event as a function of its serial position.

\textbf{Temporal clustering.} Temporal clustering measures the extent to which participants group their recall responses according to encoding position ~\citep{PolyEtal09}. For instance, if the participant recalled each item in the presentation order, this would result in a score of 1. If the participant recalled randomly with respect to presentation order, this would result in a score of ~.5.  For each event transition (and separately for each participant), we computed the rank similarity (euclidean distance) between the presentation position  of the current and next recall events. The scores were then averaged within participant to get a single number representing the extent of temporal clustering exhibited by a given participant.

\textbf{Semantic clustering.} Similar to temporal clustering, semantic clustering measures the extent to which participants group their recall responses according to semantic similarity ~\citep{PolyEtal09}. Here, we are using the topic vectors for each event as a proxy for its semantic content. Thus, similarity between the semantic content for two events can be computed by correlating their respective topic vectors.  For instance, if each consecutive recall was the next most similar event (in terms of its s), this would result in a score of ~1. If the participant recalled randomly with respect to semantic similarity, this would result in a score of ~.5.  For each event transition (and separately for each participant), we computed the rank similarity (correlation distance) between the current recall event and the next recall event. The scores were then averaged within participant to get a single number representing the extent of semantic clustering exhibited by a given participant.

@@

\begin{figure}[tb]
\centering
\includegraphics[width=1\textwidth]{figs/list_learning}
\caption{\small \textbf{Naturalistic extensions of classic list-learning memory analyses.} A). The probability of first recall as a function of the serial position of the event during encoding. B). A lag-conditional response probability curve. Given recall of event i, the probability that the next recalled item will be from serial position i +/- lag. C). Proportion of events recalled as a function of serial position. All error bars are the standard error of the mean derived from a bootstrap resampling procedure.}
\label{fig:list-learning}
\end{figure}

>>>>

\section*{Additional details about topic modeling methods and results}
\subsection*{Optimizing topic model parameters}
In order to create accurate video and recall models, we used an optimization method that was driven by our ability to explain hand-annotated memory performance metrics collected by \cite{ChenEtal17}.  Specifically, we used a grid search to compute the $\omega$ (movie sliding window duration, in scenes), $\rho$ (recall sliding window duration, in sentences), and $K$ (number of topics) that satisfied
\[
\argmax_{\omega, \rho, K} \left[\mathrm{corr}\left(\mathrm{corr}\left(\mu\left(\omega, \rho, K\right), \nu\left(\omega, \rho, K\right)\right), \theta\right)\right],
\]
where $\mathrm{corr}(\mu, \nu)$ is the per-participant correlation between the upper triangles of the temporal correlation matrices of the video ($\mu$) and recall ($\nu$) trajectory, and $\theta$ is the per-participant hand-annotated memory performance.  We searched over a grid of pre-specified values for each of these parameters; the resulting correlations are displayed in Figure~\ref{fig:paramsearch}.  The optimal parameters were $\omega = 250$, $\rho = 50$, and $K = 100$.

@@

\begin{figure}[tb]
\centering
\includegraphics[width=1\textwidth]{figs/feature_value}
\caption{\small \textbf{Feature importance analysis.} \textbf{A.} Contributions of each feature type to the structure of the video trajectory. The bar heights reflect the correlation between the video trajectory computed using all features with a video trajectory computed using all features except the indicated feature.  (Lower bars reflect features that contribute more substantially to the video trajectory's shape.) \textbf{B.} Which features are preserved during recall?  The bar heights reflect the (average) across-participant correlations between the video and recall trajectories.  Error bars denote bootstrap-estimated standard error of the mean.  \textbf{C.} Feature correlation matrix.  Each entry displays the correlation between video topic trajectories created using only the indicated (row/column) features.}
\label{fig:feature-importance}
\end{figure}

@@

@@

\subsection*{Naturalistic extensions of classic list-learning analyses}
In traditional list-learning experiments, participants view a list of items (e.g., words) and then recall the items later.  Our video-recall event matching approach affords us the ability to analyze memory in a similar way. The video and recall events can be treated analogously to studied and recalled ``items'' in a list-learning study.  We can then extend classic analyses of memory performance and dynamics (originally designed for list-learning experiments) to the more naturalistic video recall task used in our study.

Perhaps the simplest and most widely used measure of memory performance is \textit{accuracy}-- i.e., the proportion of studied (experienced) items (in this case, the 34 video events) that the participant later remembered.  \cite{ChenEtal17} developed a human rating system whereby the quality of each participant's memory was evaluated by an independent rater.  We found a strong across-participants correlation between these independant ratings and the overall number of events that our HMM approach identified in participants' recalls ($r = 0.67, p = 0.003$).

As described below, we next considered three more naunced measures of the memory performance and dynamics that are typically associated with list-learning studies.  We also provide a software package, \texttt{Quail}, for carrying out these analyses~\citep{HeusEtal17b}.

\begin{figure}[tb]
\centering
\includegraphics[width=1\textwidth]{figs/list_learning}
\caption{\small \textbf{Naturalistic extensions of classic list-learning memory analyses.} \textbf{A.} The probability of first recall as a function of the serial position of the event in the video. \textbf{B}.  The probability of recalling each event, conditioned on having just recalled the event \textit{lag} events away in the video.  \textbf{C.} The proportion of participants who recalled each event, as a function of the serial position of the events in the video.  All panels: error bars denote bootstrap-estimated standard error of the mean.}
\label{fig:list-learning}
\end{figure}

\paragraph{Probability of first recall (PFR).}  PFR curves~\citep{WelcBurn24, PostPhil65, AtkiShif68} reflect the probability that an item will be recalled first as a function of its serial position during encoding. To carry out this analysis, we initialized a number-of-participants (17) by number-of-video-events (34) matrix. Then for each participant, we found the index of the video event that was recalled first (i.e., the video event whose topic vector was most strongly correlated with that of the first recall event) and filled in that index in the matrix with a 1.  Finally, we averaged over the rows of the matrix, resulting in a 1 by 34 array representing the proportion of participant that recalled an event as a function of serial position during encoding (Fig.~\ref{fig:list-learning}A).

\paragraph{Lag conditional probability curve (lag-CRP).} The lag-CRP curve~\citep{Kaha96} reflects the probability of recalling a given event after the just-recalled event, as a function of their relative positions (or \textit{lag}).  In other words, a lag of 1 indicates that a recalled event came immediately after the previously recalled event in the video, and a lag of -3 indicates that a recalled event came 3 items before the previously recalled event.  For each recall transition (following the first recall), we computed the lag between the current recall event and the next recall event, normalizing by the total number of possible transitions.  This yielded a number-of-participants (17) by number-of-lags (-33 to +33; 67 lags total) matrix. We averaged over the rows of this matrix to obtain a group-averaged lag-CRP curve (Fig.~\ref{fig:list-learning}B).

\paragraph{Serial position curve (SPC).} SPCs~\citep{Murd62a} reflect the proportion of participants that remember each item as a function of their serial position during encoding. We initialized a number-of-participants (17) by number-of-video-events (34) matrix of zeros. Then, for each recalled event (and each participant), we found the index of the video event that the recalled event most closely matched (via the correlation between the events' topic vectors) and entered a 1 into that position in the matrix (i.e., for the given participant and event). This resulted in a matrix whose entries indicated whether or not each event was recalled by each participant (depending on whether the corresponding entires were set to one or zero.  Finally, we averaged over the rows of the matrix to yield a 1 by 34 array representing the proportion of participants that recalled each event as a function of its serial position (Fig.~\ref{fig:list-learning}C).

\paragraph{Temporal clustering scores.} Temporal clustering refers to the extent to which participants group their recall responses according to encoding position~\citep{PolyEtal09}. For instance, if the participant recalled the video events in the exact order they occurred, this would yield a score of 1.  If the participant recalled the video events in reverse order, this would yield a score of 0.  Similarly, if the participant recalled the events in random order, this yield yield an expected score of 0.5.  For each recall event transition (and separately for each participant), we sorted all not-yet-recalled events according to their absolute lag (i.e., distance away in the video).  We then computed the percentile rank of the next event the participant recalled.  We averaged these percentile ranks across all of the participant's recalls to obtain a single temporal clustering score for the participant.  Overall, we found that participants with higher temporal clustering scores also tended to recall more events ($r = 0.62, p = 0.007$).

\paragraph{Semantic clustering scores.} Semantic clustering measures the extent to which participants grouped their recall responses according to semantic similarity~\citep{PolyEtal09}. Here, we used the topic vectors for each event as a proxy for its semantic content. Thus, the similarity between the semantic content for two events can be computed by correlating their respective topic vectors.  For each recall event transition, we sorted all not-yet-recalled events according to how correlated the topic vector \textit{of the closest-matching video event} was to the topic vector of the closest-matching video event to the just-recalled event.  We then computed the percentile rank of the observed next recall.  We averaged these percentile ranks across all of the participant's recalls to obtain a single semantic clustering score for the participant.  We found that participants who exhibited stronger semantic clustering scores overall remembered more video events ($r = 0.55, p = 0.02$).

\textbf{JRM STOPPED HERE}

\subsection*{Additional measures of naturalistic memory}
To quantify the similarity between the video model and individual recall models, we considered a number of novel metrics.  First, we tested whether each participant's recall model matched the video model in a general sense. To do this, for each participant we filtered the video model to only include the events that the participant remembered and computed the root mean squared difference (RMSD) between the video model and the recall model. As an example, if the participant remembered all the events in order (with perfect precision), the expected distance value would be 0. However, if they remembered a subset of events, events out of order, or with low precision, the expected distance would be greater than 0. To assess significance, we performed a permutation test where we circularly shifted the recall model (10000 times) and recomputed the RMSD. The recall model significantly matched the video model for nine of the participants ($p < 0.05$, participants: 3-4, 8-13, 17 and the p-value for the rest of the participants was less than .25). Furthermore, the RMSD values were significantly correlated to hand annotated memory performance across participants (Pearson's $r(16) =-.57, p = 0.016$). Thus, a closer match between the video and recall event models was related to better recall performance.

@@

@@
